Hydras \& Co.

### 2023-10-26 doc/chapter-primrec.tex
Nat2 term coercion (#166)
* minor changes

* Making isPR a type class (to resume)

* Converting isPR into a type class

* Making isPR a class (continued)

* IsPR is now a type class
Starting to improve the presentation of primitive recursive functions

\begin{description}
  \item[zero] the constant function of value $0$ is primitive recursive.
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any pair $0< i\leq n$, the projection $\pi_{i,n}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{i,n}(x_1,x_2,\dots,x_{n})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}: \mathbb{N}^n\rightarrow\mathbb{N}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1})): \mathbb{N}^n\rightarrow\mathbb{N}$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

@@

  We present in this chapter a formalization of  primitive recursive functions, taken from  Russel O'Connor's formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}.

@@

Another example?
Let us consider the following term\footnote{Of course, we never typed this term \emph{verbatim}; we obtained it by an interactive proof the reader will be able to make after 
reading Sect.\vref{sect:proofs-of-isPR}.}:

\label{sect:bigfac}

\input{movies/snippets/PrimRecExamples/bigPRa}

Let us now interpret this term as an arithmetic function.

\input{movies/snippets/PrimRecExamples/bigPRb}

After this test, the term \texttt{bigPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections simple ways to prove that a given function is primitive recursive, without building such an unreadable term.

@@

\input{movies/snippets/PrimRecExamples/isPRExtEqualTrans}

>>>>

\begin{description}
  \item[zero] the natural number $0$ is a primitive recursive function without arguments (in other words, a  \emph{constant}, or a \emph{nullary} function).
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any $i$ and $n$, such that $0< i\leq n$, the projection $\pi_{i,n}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{i,n}(x_1,x_2,\dots,x_{n})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}: \mathbb{N}^n\rightarrow\mathbb{N}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1})): \mathbb{N}^n\rightarrow\mathbb{N}$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

@@

\subsection{A few (informal) examples}
Before playing with primitive recursive functions in \coq{}, let us get familiar with the definition above, with the help of a few simple examples.
\label{sect:primrec-examples1}

\subsubsection{Projections}

For instance, the projection $\pi{2,3}$ satisfies the equation
$\pi{1,3}(x,y,z)=y$ for any natural numbers $x$, $y$ and $z$.

\subsubsection{The constant function of value 0}
\label{sect:k0}

The \emph{unary} function which always returns $0$ may be defined through the \emph{composition} construction (with $n=1$, $m=0$, and $h=\texttt{zero}$).

\subsection{Addition on natural natural numbers}

Addition may be defined by primitive recursion:

\begin{align}
  0 + x_1 &:= x_1 \\
  S x_0 + x_1 &:= S (x_0 + x_1)
\end{align}

Both equations can be rewritten as follows:

\begin{align}
0 + x_1 &:= g(x_1) \\
S p + x_1 &:= h(p, p+x_1, x_1) \\
\textbf{where}\; g(x_1)&:= x_1 \\
\textbf{and}\; h(p,x, x_1) &:= S\;x 
\end{align}

It remains to show that $g$ and $h$ are primitive recursive, which is almost immediate:

\begin{itemize}
\item $g$ is the projection $\pi_{1,1}$,
\item $h$ is the composition (with $n=3$ and $m=1$) of the function \textbf{S} and the projection $\pi_{2,3}$ .
\end{itemize}

\subsection{Multiplication on natural natural numbers}

The following equations define the product of two natural numbers:

\begin{align}
  0 \times x_1 &:= 0 \\
  S x_0 \times x_1 &:= (x_0 \times x_1) + x_1
\end{align}

This function is an instance of the primitive recursion scheme,
with $n=1$, $g$ the constant unary function which returns $0$ (see Subsection~\vref{sect:k0}),
 and $h$ the function defined by $h(p,x,n_1)=x+n_1$, which can be written as the composition of $+$ and the projections 
$\pi_{2,3}$ and $\pi_{3,3}$ (The three of them  being  primitive recursive).

\index{hydras}{Exercises}
\begin{exercise}
 Show that the factorial function is primitive recursive (by the same method as above). 
\end{exercise}

\begin{remark}

The four examples above are defined in \coq in Section~\vref{sect:primrec-examples2}

\end{remark}
  We present in this chapter a formalization of  primitive recursive functions, taken from  Russel O'Connor's formalization in \coq{} of
G\"odel's incompleteness theorems~\cite{OConnor05}.

@@

\subsection{A few arithmetic functions}
\label{sect:primrec-examples2}
Let us show how the few functions presented in Section~\vref{sect:primrec-examples2} can be defined in \coq.

\subsubsection{Constant function of value 0}

\inputsnippets{PrimRecExamples/cst0}

\subsubsection{Addition}

\inputsnippets{PrimRecExamples/addition}

\subsubsection{Multiplication}

\inputsnippets{PrimRecExamples/multiplication}

\subsubsection{The factorial function}

\inputsnippets{PrimRecExamples/factorial}

@@

\input{movies/snippets/primRec/isPRExtEqual}
